Followers Are Not Enough: Beyond Structural Communities in Online Social Networks

### 2014-04-30 abstract-eccs14/eccs14_abstract.tex
Beginning to incorporate reviewer comments.

Community detection in online social networks is typically based on the analysis of the explicit connections between users, such as ``friends" on Facebook and ``followers" on Twitter. This so-called \emph{structure-based} approach, relies solely on the reported relationships between users---completely ignoring the users activity, content, or interactions. However, users often have hundreds or even thousands of such structural-connections, many of which do not correspond to real friendships or even accounts the user interacts with. We claim that community detection in online social networks should be \emph{question-oriented} and rely on additional information beyond the simple (often spurious) structure of the network. The concept of `community' is very general, and we have found that different questions such as ``who does a user interact with?", ``from whom does a user receive information?" and ``with whom does a user share similar interests?" can lead to the discovery of many rich \emph{and} interesting communities that the structure-based analysis alone is completely blind to. 
%We propose looking at when and how communities motivated by these different questions overlap, and whether different approaches to asking the question, ``What community are you in?'' leads to different insights about a social network. 
For example, a user on Twitter might connect mostly with computational social scientists, talk mostly about machine learning, interact solely with close friends (who may or may not be computational social scientists), and utilize the service only on nights and weekends. We argue that each of these different `profiles' of the user highlight different views of the user's social network, and represent different types of communities.

We divide our approaches into four categories based on the questions outlined above: structure-based, activity-based, topic-based, and interaction-based. The structure-based approach, outlined above, relies solely on the known follower relationships, and is used here purely for comparison with the standard. To detect the \emph{question-oriented} communities we weight the structural edges with different scores derived to both answer and give insight into the question posed about the network. For the \emph{activity-based} communities, we consider only the timing of each user's tweets and ignore any additional content, i.e., we view the behavior of a user $u$ on Twitter as a point process, where at any instant $t$ the user has either emitted a tweet ($X_{t}(u) = 1$) or remained silent ($X_{t}(u) = 0$)---reducing all of the information generated by a user to a symbolic time series: $\{ X_{t}(u)\}$. We then compute the directed flow of information between two users $u$ and $v$ by computing the so-called \emph{transfer entropy} between their time series $X_{t}(u)$ and $X_{t}(v)$; a novel approach which gives interesting insights into how information transmits in a network as well as giving insight into who the influential users are. For the \emph{interaction-based} communities, we weight the structural network with a measure proportional to the number of mentions and/or retweets between users. Lastly, in order to detect the topical communities, we weight the edges of the structural network through a measure based on the number of common hashtags between pairs of users. Hashtags are a good proxy for a tweet's content as hashtags are explicitly meant to be keywords indicating the topic of the tweet. It should be noted that a user may be a member of multiple communities simultaneously in a given network, e.g., a user may tweet about ``Complexity" \emph{and} "Lucca, Italy", which may be two separate topical communities, but may overlap for example when a collection of users tweet about ECCS'14. For this reason we seek overlapping communities---often called \emph{coverings}.

To detect covering of each weighted, directed network we use the OSLOM algorithm~\cite{LancichinettiPlos} . We then applied several statistical measures (e.g., mutual information) to compare the communities derived from the different questions with intriguing and informative results.  Not surprisingly, there was similarity between coverings dictated by each generic community type (structural, activity-based, etc.).
% e.g., the coverings derived from transfer-entropy are more similar to each other than to any of the other coverings, with a similar result for the mention, retweet, and mention-retweet coverings. 
Interestingly, the coverings resulting from the \emph{question-oriented} weightings were all more similar to each other than to the covering from the \emph{structure-based} analysis. We found that the coverings derived from the \emph{topic-based} questions (hashtag weightings) were significantly different from all of the other question-oriented coverings.

>>>>

Community detection in online social networks is typically based on the analysis of the explicit connections between users, such as ``friends" on Facebook and ``followers" on Twitter. This so-called \emph{structure-based} approach relies solely on the reported relationships between users---completely ignoring the users activity, content, or interactions. However, users often have hundreds or even thousands of such structural-connections, many of which do not correspond to real friendships or even accounts the user interacts with. We claim that community detection in online social networks should be \emph{question-oriented} and rely on additional information beyond the simple (often spurious) structure of the network. The concept of `community' is very general, and we have found that different questions such as ``who does a user interact with?", ``from whom does a user receive information?" and ``with whom does a user share similar interests?" can lead to the discovery of many rich \emph{and} interesting communities that the structure-based analysis alone is completely blind to. 
%We propose looking at when and how communities motivated by these different questions overlap, and whether different approaches to asking the question, ``What community are you in?'' leads to different insights about a social network. 
For example, a user on Twitter might connect mostly with computational social scientists, talk mostly about machine learning, interact solely with close friends (who may or may not be computational social scientists), and utilize the service only on nights and weekends. We argue that each of these different `profiles' of the user highlight different views of the user's social network, and represent different types of communities.

We divide our approaches into four categories based on the questions outlined above: structure-based, activity-based, topic-based, and interaction-based. The structure-based approach, outlined above, relies solely on the known follower relationships, and is used here purely for comparison with the standard. To detect the \emph{question-oriented} communities we weight the structural edges with different scores derived to both answer and give insight into the question posed about the network. For the \emph{activity-based} communities, we consider only the timing of each user's tweets and ignore any additional content, i.e., we view the behavior of a user $u$ on Twitter as a point process, where at any instant $t$ the user has either emitted a tweet ($X_{t}(u) = 1$) or remained silent ($X_{t}(u) = 0$)---reducing all of the information generated by a user to a symbolic time series: $\{ X_{t}(u)\}$. We then compute the directed flow of information between two users $u$ and $v$ by computing the \emph{transfer entropy} between their time series $X_{t}(u)$ and $X_{t}(v)$; an information theoretic approach which gives insights into how information transmits in a network as well as insights into who the influential users are. For the \emph{interaction-based} communities, we weight the structural network with a measure proportional to the number of mentions and/or retweets between users. Lastly, in order to detect the topical communities, we weight the edges of the structural network through a measure based on the number of common hashtags between pairs of users. Hashtags are a good proxy for a tweet's content as hashtags are explicitly meant to be keywords indicating the topic of the tweet. It should be noted that a user may be a member of multiple communities simultaneously in a given network, e.g., a user may tweet about ``Complexity'' \emph{and} ``Lucca, Italy'', which may be two separate topical communities, but may overlap for example when a collection of users tweet about ECCS'14. For this reason we seek overlapping communities---often called \emph{coverings}.

To detect covering of each weighted, directed network we use the OSLOM algorithm~\cite{LancichinettiPlos}. We then applied several statistical measures (e.g., mutual information) to compare the communities derived from the different questions with intriguing and informative results.  Not surprisingly, there was similarity between coverings dictated by each generic community type (structural, activity-based, etc.).
% e.g., the coverings derived from transfer-entropy are more similar to each other than to any of the other coverings, with a similar result for the mention, retweet, and mention-retweet coverings. 
Interestingly, the coverings resulting from the \emph{question-oriented} weightings were all more similar to each other than to the covering from the \emph{structure-based} analysis. We found that the coverings derived from the \emph{topic-based} questions (hashtag weightings) were significantly different from all of the other question-oriented coverings.

### 2014-04-30 paper-for_arxiv/intro.tex
Beginning to incorporate reviewer comments.

The activity-based, topic-based, and interaction-based networks allow us to build a more complete picture of the \emph{latent} social network present in online social media, as opposed to the \emph{explicit} social network indicated by structural links. In this paper, we explore the relation between these various possible networks and their corresponding communities. We begin by describing the methodologies used to generate the various types of networks, and infer their community structure. We then explore how the communities of users differ depending on the type of network used. Finally, we explore how communication patterns differ across and within the different community types.

>>>>

The activity-based, topic-based, and interaction-based networks allow us to build a more complete picture of the \emph{implicit} social network present in online social media, as opposed to the \emph{explicit} social network indicated by structural links. In this paper, we explore the relation between these various possible networks and their corresponding communities. We begin by describing the methodologies used to generate the various types of networks, and infer their community structure. We then explore how the communities of users differ depending on the type of network used. Finally, we explore how communication patterns differ across and within the different community types.

### 2014-04-30 paper-for_arxiv/methods.tex
Beginning to incorporate reviewer comments.

Let $\{X_{t}\}$ and $\{ Y_{t}\}$ be two strong-sense stationary stochastic processes. We use the notation $X_{t-k}^{t}$ to denote the values of the stochastic process from time $t-k$ to time $t$, $X_{t-k}^{t} = (X_{t-k}, X_{t-(k-1)}, \ldots, X_{t - 1}, X_{t})$. The lag-$k$ transfer entropy of $Y$ on $X$ is defined as 
\begin{align}
	\text{TE}_{Y \to X}^{(k)} &= H\left[X_{t} | X_{t-k}^{t-1}\right] - H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right],
\end{align}
where
\begin{align}
	H\left[X_{t} | X_{t-k}^{t-1}\right] = - E\left[ \log_{2} p(X_{t} | X_{t-k}^{t-1}) \right]
\end{align}
and 
\begin{align}
	H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right] = - E\left[ \log_{2} p(X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}) \right]
\end{align}
are the usual conditional entropies over the conditional (predictive) distributions $p(x_{t} | x_{t-k}^{t-1})$ and $p(x_{t} | x_{t-k}^{t-1}, y_{t-k}^{t-1})$. This formulation was originally developed in~\cite{schreiber2000measuring}, where transfer entropy was proposed as an information theoretic measure of \emph{directed} information flow. Formally, recalling that $H\left[X_{t} | X_{t-k}^{t-1}\right]$ is the uncertainty in $X_{t}$ given its values at the previous $k$ time points, and that $H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]$ is the uncertainty in $X_{t}$ given the joint process $\{(X_{t}, Y_{t})\}$ at the previous $k$ time points, transfer entropy measures the reduction in uncertainty of $X_{t}$ by including information about $Y_{t-k}^{t-1},$ controlling for the information in $X_{t - k}^{t-1}$. By the `conditioning reduces entropy' result~\cite{cover2012elements}
\begin{align}
	H[X | Y, Z] \leq H[X | Y],
\end{align}
we can see that transfer entropy is always non-negative, and is zero precisely when $H\left[X_{t} | X_{t-k}^{t-1}\right] = H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]$, in which case knowing the past $k$ lags of $Y_{t}$ does not reduce the uncertainty in $X_{t}$. If the transfer entropy is positive, then $\{ Y_{t}\}$ is considered causal for $\{ X_{t}\}$ in the Granger sense~\cite{granger1963economic}.
% More generally we can define the transfer entropy as the limit of lag-$k$ transfer entropies,
% \begin{align}
% 	\text{TE}_{Y \to X} &= \lim_{k \to \infty} \text{TE}_{Y \to X}^{(k)},
% \end{align}
% if the limit exists.

In estimating the transfer entropy from finite data, we will assume that the process $(X_{t}, Y_{t})$ is jointly stationary, which gives us that
\begin{align}
	p(x_{t} | x_{t-k}^{t-1}) = p(x_{k+1} | x_{1}^{k})
\end{align}
and
\begin{align}
	p(x_{t} | x_{t-k}^{t-1}, y_{t-k}^{t-1}) = p(x_{k+1} | x_{1}^{k}, y_{1}^{k})
\end{align}
for all $t$. That is, the predictive distribution only depends on the past, not on when the past is observed\footnote{We really only need \emph{conditional} stationarity~\cite{caires2003nonparametric}, but stationarity implies conditional stationarity.}. Given this assumption, we compute estimators for $p(x_{k+1} | x_{1}^{k})$ and $p(x_{k+1} | x_{1}^{k}, y_{1}^{k})$ by `counting': for each possible past $(x_{1}^{k}, y_{1}^{k})$, we count the number of times a future of type $x_{k+1}$ occurs, and normalize. Call these estimators $\hat{p}(x_{k+1} | x_{1}^{k})$ and $\hat{p}(x_{k+1} | x_{1}^{k}, y_{1}^{k})$. Then the plug-in estimator for the transfer entropy is
\begin{align}
	\widehat{\text{TE}}_{Y \to X}^{(k)} &= \hat{H}\left[X_{t} | X_{t-k}^{t-1}\right] - \hat{H}\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]
\end{align}
where we use the plug-in estimators $\hat{H}\left[X_{t} | X_{t-k}^{t-1}\right]$ and $\hat{H}\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]$ for the entropies. It is well known that the plug-in estimator for entropy is biased~\cite{paninski2003estimation}. To account for this bias, we use the Miller-Madow adjustment to the plug-in estimator~\cite{miller1955note}.

>>>>

Let $\{X_{t}\}$ and $\{ Y_{t}\}$ be two strong-sense stationary stochastic processes\footnote{Recall that a a stochastic process is strong-sense stationary if the distribution over process evaluated at finitely many time points is invariant to an overall timeshift~\cite{grimmett2001probability}.}. In our work, these would correspond to two user activities, $X_{t}(u)$ and $X_{t}(v).$ We use the notation $X_{t-k}^{t}$ to denote the values of the stochastic process from time $t-k$ to time $t$, $X_{t-k}^{t} = (X_{t-k}, X_{t-(k-1)}, \ldots, X_{t - 1}, X_{t})$. The lag-$k$ transfer entropy of $Y$ on $X$ is defined as 
\begin{align}
	\text{TE}_{Y \to X}^{(k)} &= H\left[X_{t} | X_{t-k}^{t-1}\right] - H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right],
\end{align}
where
\begin{align}
	H\left[X_{t} | X_{t-k}^{t-1}\right] = - E\left[ \log_{2} p(X_{t} | X_{t-k}^{t-1}) \right]
\end{align}
and 
\begin{align}
	H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right] = - E\left[ \log_{2} p(X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}) \right]
\end{align}
are the usual conditional entropies over the conditional (predictive) distributions $p(x_{t} | x_{t-k}^{t-1})$ and $p(x_{t} | x_{t-k}^{t-1}, y_{t-k}^{t-1})$. This formulation was originally developed in~\cite{schreiber2000measuring}, where transfer entropy was proposed as an information theoretic measure of \emph{directed} information flow. Formally, recalling that $H\left[X_{t} | X_{t-k}^{t-1}\right]$ is the uncertainty in $X_{t}$ given its values at the previous $k$ time points, and that $H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]$ is the uncertainty in $X_{t}$ given the joint process $\{(X_{t}, Y_{t})\}$ at the previous $k$ time points, transfer entropy measures the reduction in uncertainty of $X_{t}$ by including information about $Y_{t-k}^{t-1},$ controlling for the information in $X_{t - k}^{t-1}$. By the `conditioning reduces entropy' result~\cite{cover2012elements}
\begin{align}
	H[X | Y, Z] \leq H[X | Y],
\end{align}
we can see that transfer entropy is always non-negative, and is zero precisely when $H\left[X_{t} | X_{t-k}^{t-1}\right] = H\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]$, in which case knowing the past $k$ lags of $Y_{t}$ does not reduce the uncertainty in $X_{t}$. If the transfer entropy is positive, then $\{ Y_{t}\}$ is considered causal for $\{ X_{t}\}$ in the Granger sense~\cite{granger1963economic}.
% More generally we can define the transfer entropy as the limit of lag-$k$ transfer entropies,
% \begin{align}
% 	\text{TE}_{Y \to X} &= \lim_{k \to \infty} \text{TE}_{Y \to X}^{(k)},
% \end{align}
% if the limit exists.

In estimating the transfer entropy from finite data, we will assume that the process $(X_{t}, Y_{t})$ is jointly stationary, which gives us that
\begin{align}
	p(x_{t} | x_{t-k}^{t-1}) = p(x_{k+1} | x_{1}^{k})
\end{align}
and
\begin{align}
	p(x_{t} | x_{t-k}^{t-1}, y_{t-k}^{t-1}) = p(x_{k+1} | x_{1}^{k}, y_{1}^{k})
\end{align}
for all $t$. That is, the predictive distribution only depends on the past, not on when the past is observed. Given this assumption, we compute estimators for $p(x_{k+1} | x_{1}^{k})$ and $p(x_{k+1} | x_{1}^{k}, y_{1}^{k})$ by `counting': for each possible past $(x_{1}^{k}, y_{1}^{k})$, we count the number of times a future of type $x_{k+1}$ occurs, and normalize. Call these estimators $\hat{p}(x_{k+1} | x_{1}^{k})$ and $\hat{p}(x_{k+1} | x_{1}^{k}, y_{1}^{k})$. Then the plug-in estimator for the transfer entropy is
\begin{align}
	\widehat{\text{TE}}_{Y \to X}^{(k)} &= \hat{H}\left[X_{t} | X_{t-k}^{t-1}\right] - \hat{H}\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]
\end{align}
where we use the plug-in estimators $\hat{H}\left[X_{t} | X_{t-k}^{t-1}\right]$ and $\hat{H}\left[X_{t} | X_{t-k}^{t-1}, Y_{t-k}^{t-1}\right]$ for the entropies. It is well known that the plug-in estimator for entropy is biased~\cite{paninski2003estimation}. To account for this bias, we use the Miller-Madow adjustment to the plug-in estimator~\cite{miller1955note}.

### 2014-04-30 paper-for_arxiv/results.tex
Beginning to incorporate reviewer comments.

Each \emph{column} of Figure~\ref{Fig-distributions_by_types} demonstrates how the distribution of weights change with the community type used to induce the partition. For example, the second column shows how the distributions of hashtag weights change based on the community partition. Unsurprisingly, we see that the greatest difference between the distributions occurs when we consider the topic-based communities, since the topic-based communities are determined based on the hashtag weights. However, we also see differences in the distributions when the Community-type / Weight-type pairs do not match. For example, for the covering corresponding to the activity-based community, the distribution of

\begin{figure}[!h]
	\centering
	\includegraphics[width=1\textwidth]{distributions-edges.eps}
	\caption{The distribution of edge weights for different community types and weight types. }
	\label{Fig-distributions_by_types}
\end{figure}

>>>>

Each \emph{column} of Figure~\ref{Fig-distributions_by_types} demonstrates how the distribution of weights change with the community type used to induce the partition. For example, the second column shows how the distributions of hashtag weights change based on the community partition. Unsurprisingly, we see that the greatest difference between the distributions occurs when we consider the topic-based communities, since the topic-based communities are determined based on the hashtag weights. However, we also see differences in the distributions when the Community-type / Weight-type pairs do not match. For example, for the covering corresponding to the activity-based communities, the hashtag weights tend to be higher for edges internal to communities than between communities. Similarly, for the covering corresponding to the topic-based communities, the transfer entropy weights tend to be higher on edges within (intra) and between (inter) communities, and lower for edges between members with multiple, non-identical memberships (multi).

\begin{figure}[!h]
	\centering
	\includegraphics[width=1\textwidth]{distributions-edges.eps}
	\caption{The distribution of edge weights for different community types (rows) and weight types (columns). The red,  blue, and magenta values below each distribution indicate the median of weight on intra-, inter-, and multi-edges, respectively.}
	\label{Fig-distributions_by_types}
\end{figure}
